Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-15 write-ups/PLDI/paper.tex
Remove figure of overhead

The example of \myfig\ref{fig:cgraphsearch} is simpler than, but representative of,
most level-by-level graph algorithms.  These algorithms tend to
require some state to be transmitted, and in some cases
a global barrier is required in the midst of a transaction, in which
case $\globalbarrier$, but not $\resizingglobalbarrier$, is suitable
for achieving synchronization without disturbing the set of active
workgroups.  But often it is the case that on completing a level of
the graph algorithm, the next level could be processed by more or
fewer workgroups, something which $\resizingglobalbarrier$
facilitates.

@@

\begin{table}
\small
\centering
\begin{tabular}{ l c c c c c c}
App. & resizing barr. & kill & fork & transmit & LoC & data sets\\
\hline
\rowcolor{Gray1}
color & 2 / 2 & 0 & 0 & 4 & 55 & 2\\
\rowcolor{Gray1}
mis & 3 / 3 & 0 & 0 & 0 & 71 & 2\\
\rowcolor{Gray1}
bc & 3 / 6 & 0 & 0 & 3 & 150 & 2\\
\rowcolor{Gray1}
p-sssp & 3 / 3 & 0 & 0 & 0  & 42 & 1\\
\rowcolor{Gray2}
bfs & 2 / 2 & 0 & 0  & 4  & 185 & 2\\
\rowcolor{Gray2}
l-sssp & 2 / 2 & 0 & 0  & 4  & 196 & 2\\
\rowcolor{Gray3}
octree & 0 / 0 & 1 & 1 & 0 & 213 & 1 \\
\rowcolor{Gray3}
game & 0 / 0 & 1 & 1 & 0 & 308 & 1 \\
\end{tabular} \\
\vspace{.2cm}
\crule[Gray1]{.2cm}{.2cm} Pannotia \hspace{.4cm} \crule[Gray2]{.2cm}{.2cm} Lonestar GPU  \hspace{.4cm}  \crule[Gray3]{.2cm}{.2cm} work stealing
\caption{Blocking GPU applications investigated}
\label{tab:applications}
\end{table}

\ADComment{I thought we had some examples where we actually wanted a non-resizing barrier; we say this earlier in the paper so we need to smooth.}

\ADComment{Missing: a discussion of what we did not port.}

@@

@@

\ADComment{Check:} Among the GPUs available to us, four met our
requirements: Intel HD 520, HD 5500 and Iris 6100, and AMD Radeon R7.
However, AMD's Linux drivers for OpenCL suffer from a known defect
whereby long-running kernels lead to defunct processes that the OS
cannot kill.  We observed this on our AMD platform with the latest
driver, and simultaneously obtained alarming results on this platform
when using SVM atomics: modifying a kernel to include a simple
host/device handshake through SVM slowed down kernel execution by an
order of magnitude in some cases.  This behaviour, combined with the
defunt processes issue, lead us to suspect that the support we require
for cooperative kernels is not yet mature enough within AMD's drivers,
which we plan to discuss with engineers at AMD.

@@

As a proxy for genuine graphics rendering, we devised the following
procedure to determine non-cooperative OpenCL workloads that simulate
the computational intensity of various graphics rendering workloads.
%
\ADComment{I think this is good but could be clearer and it would be helpful to discuss.}
We repetitively launch a kernel with a
controlled workload and detect from which workload size the screen
animation starts to jitter. \ADComment{Be clear that this is manual.} Given the kernel workload and its
scheduling frequency, we can deduce what is the workload and frequency
needed by the system to maintain a smooth animation. We use this
approach to measure the system workloads needed to maintain smooth
animations of three types: (\emph{light}) emphasis of desktop icons
under the mouse pointer, (\emph{medium}) window dragging over the
desktop, and (\emph{heavy}) WebGL animations in a browser.

\begin{table}
\small
\centering
\begin{tabular}{ l r r }
Workload & Duration (ms) & Period (ms)\\
\hline
light & 3 & 70\\
medium & 3 & 40\\
heavy & 10 & 40\\
\end{tabular}
\caption{\ADComment{Clearer caption?} Characteristics of non-cooperative kernels workload}
\label{tab:noncooperativeworkload}
\end{table}

\paragraph{Results}
\mytab~\ref{tab:noncooperativeworkload} illustrates the caracteristics
found for the three types of workloads. For the medium and heavy
workloads, the 40ms period corresponds to an execution every $25^{th}$
of seconds, which coincides with the human persistence of vision. The
3ms execution duration of both light and medium configurations may
indicate a lower bound of GPU execution time for system workloads.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{cooperative_overhead.pdf}
\caption{Overhead: ratio of execution time}\label{fig:overhead}
\end{figure}

@@

\paragraph{Results}
For each GPU, \mytab\ref{tab:overhead} shows the geometric mean and
maximum slowdown across all applications and data sets, each benchmark
begin run 10 times. For the maximum slowdowns, we indicate which
application and data set was responsible.  The slowdown is below 2 even
in the worst case, and closer to 1 on average. We consider these results
as encouraging, especially since the performance of our prototype could
clearly be improved upon if cooperative scheduling was implemented
directly in the GPU driver.

>>>>

The example of \myfig\ref{fig:cgraphsearch} is simpler than, but representative of,
most level-by-level graph algorithms.
It is typically the case that on completing a level of
the graph algorithm, the next level could be processed by more or
fewer workgroups, something which $\resizingglobalbarrier$
facilitates.  As in \myfig\ref{fig:cgraphsearch}, some level-specific state must be transmitted to new workgroups.

@@

\begin{table}
\small
\centering
\begin{tabular}{ l r r r r r r}
App. & resizing barr. & kill & fork & transmit & LoC & data sets\\
\hline
\rowcolor{Gray1}
color & 2 / 2 & 0 & 0 & 4 & 55 & 2\\
\rowcolor{Gray1}
mis & 3 / 3 & 0 & 0 & 0 & 71 & 2\\
\rowcolor{Gray1}
bc & 3 / 6 & 0 & 0 & 3 & 150 & 2\\
\rowcolor{Gray1}
p-sssp & 3 / 3 & 0 & 0 & 0  & 42 & 1\\
\rowcolor{Gray2}
bfs & 2 / 2 & 0 & 0  & 4  & 185 & 2\\
\rowcolor{Gray2}
l-sssp & 2 / 2 & 0 & 0  & 4  & 196 & 2\\
\rowcolor{Gray3}
octree & 0 / 0 & 1 & 1 & 0 & 213 & 1 \\
\rowcolor{Gray3}
game & 0 / 0 & 1 & 1 & 0 & 308 & 1 \\
\end{tabular} \\
\vspace{.2cm}
\crule[Gray1]{.2cm}{.2cm} Pannotia \hspace{.4cm} \crule[Gray2]{.2cm}{.2cm} Lonestar GPU  \hspace{.4cm}  \crule[Gray3]{.2cm}{.2cm} work stealing
\caption{Blocking GPU applications investigated}
\label{tab:applications}
\end{table}

@@

From the Lonestar suite we omit two applications that use global barriers:
dmr has been shown to be unstable across GPU vendors~\cite{DBLP:conf/iwocl/SorensenD16}, and mst is trivially short-running so that it is not
interesting for our multitasking approach.
%
These exceptions aside, we have handled all the available OpenCL
implemenatinos of irregular algorithms that we are aware of, plus two
easy-to-adapt CUDA algorithms (the workstealing examples).

@@

Among the GPUs available to us, four met our
requirements: Intel HD 520, HD 5500 and Iris 6100, and AMD Radeon R7.
However, AMD's Linux drivers for OpenCL suffer from a known defect
whereby long-running kernels lead to defunct processes that the OS
cannot kill.\footnote{See \url{http://developer.amd.com/download/AMD_APP_SDK_Release_Notes_Developer.pdf}}  We observed this on our AMD platform with the latest
driver, and simultaneously obtained alarming results on this platform
when using SVM atomics: modifying a kernel to include a simple
host/device handshake through SVM slowed down kernel execution by an
order of magnitude in some cases.  This behaviour, combined with the
defunt processes issue, lead us to suspect that the support we require
for cooperative kernels is not yet mature enough within AMD's drivers,
which we plan to discuss with engineers at AMD.

@@

As a proxy for genuine graphics rendering, we devised the following
procedure to determine non-cooperative OpenCL workloads that simulate
the computational intensity of various graphics rendering workloads.
%
We designed a synthetic kernel that occupies all workgrouops of a
given GPU for a parameterised length of time $t$.  We wrote a host
program to repeatedly launch this kernel in an infinite loop.  We then
searched for a maximum value for the parameter $t$ that allowed the
synthetic kernel to execute without having an observable impact on
graphics rendering, where ``observable'' was decided manually by one
of the paper authors.  Having found the appropriate value for $t$, we
ran the host application for $X$ seconds, measuring the time $Y < X$
dedicated to GPU execution during this period, as well as the number
of kernel launches $n$ that were issued.  We used $X \geq 10$ in all
experiments.  The values $(X-Y)/n$ and $X/n$ estimate the average time
spent using the GPU to render the display between kernel calls and the
rate at which the OS requires the GPU for display rendering,
respectively.

We used this approach to measure the GPU availability required for
three types of rendering: \emph{light}, whereby desktop icons are
smoothly emphasised under the mouse pointer; \emph{medium}, whereby
window dragging over the desktop is smoothly animated; and
\emph{heavy}, which requires smooth animation of a WebGL shader in a
browser.  For \emph{heavy} we used the Paper
Planes\footnote{\url{https://www.chromeexperiments.com/experiment/paper-planes}}
and Kaleidoscope\footnote{\url{http://colordodge.com/Kaleidoscope/}}
shaders running in a recent version of Firefox.

\begin{table}
\small
\centering
\begin{tabular}{ l r r }
Workload & Duration (ms) & Rate (ms)\\
\hline
light & 3 & 70\\
medium & 3 & 40\\
heavy & 10 & 40\\
\end{tabular}
\caption{Computed estimates of the GPU requirements for three different classes of graphics rendering}
\label{tab:noncooperativeworkload}
\end{table}

\mytablong\ref{tab:noncooperativeworkload} illustrates the data
computed for the three types of workloads across our GPUs.  The
\emph{Duration} and \emph{Rate} columns report values corresponding to
$(X-Y)/n$ and $X/n$ in the above discussion, respectively.  For the
medium and heavy workloads, the 40ms rate corresponds to graphics
rendering occuring 25 times per second, which coincides with the human
persistence of vision. The 3ms execution duration of both light and
medium configurations indicates that, as expected, GPU computation is
cheaper for basic display rendering compared with rendering that uses
complex shaders.

We use these results to guide our evaluation of the responsiveness of
cooperative scheduling in \mysec\ref{sec:responsiveness}.

@@

@@

\paragraph{Results}
For each GPU, \mytab\ref{tab:overhead} shows the geometric mean and
maximum slowdown across all applications and data sets, with averages and maxima computed over 10 runs per benchmark. For the maximum slowdowns, we indicate which
application and data set was responsible.  The slowdown is below 2 even
in the worst case, and closer to 1 on average. We consider these results
as encouraging, especially since the performance of our prototype could
clearly be improved upon if cooperative scheduling was implemented
directly in the GPU driver.
